Supporting Information: Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes.

### 2020-05-13 Ben_Manuscripts/hdphmm/Supporting_Information.tex
New explanations

  \subsection{Clustering on the self-transition probabilities}\label{section:selfT_cluster}

  To cluster on $T_{ii}$, we first
  cast it in terms of the expected dwell times. The expected value of the number of sequential
  self-transitions is simply $\frac{1}{1 - T_{ii}}$. This relationship correctly implies that 
  dwell times approach infinity as $T_{ii}$ approaches 1. There we found the most success with
  agglomerative clustering by linearizing this relationship and clustering on $-log(1 - T_{ii})$.

  \subsection{Choosing the Number of Clusters}\label{section:nclusters}

  \begin{figure}[h]
  \centering
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{silhouette_MET.pdf}
  \caption{methanol}\label{fig:silhouette_MET}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{silhouette_URE.pdf}
  \caption{urea}\label{fig:silhouette_URE}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{silhouette_GCL.pdf}
  \caption{ethylene glycol}\label{fig:silhouette_GCL}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{silhouette_ACH.pdf}
  \caption{acetic acid}\label{fig:silhouette_ACH}
  \end{subfigure}
  \caption{We used the silhouette score test in order to quantitatively
  inform our decision on the numbrer of clusters. Silhouette scores 
  closer to one indicate good clustering. The crosses on the plot indicate
  our choice in number of clusters after careful consideration of the
  silhouette scores in conjunction with qualitative examination of the clusters
  formed by the choice. Although the test often
  favors fewer states, we elected to use an intermediate number of states in 
  the cases of $A$ and $\Sigma$ because choosing 2--3 clusters 
  insufficiently separates the dynamical modes. }\label{fig:silhouette}
  \end{figure}

>>>>

%  \subsection{Clustering on the self-transition probabilities}\label{section:selfT_cluster}
%  
%  To cluster on $T_{ii}$, we first cast it in terms of the expected dwell times. The expected 
%  value of the number of sequential self-transitions is simply $\frac{1}{1 - T_{ii}}$. This 
%  relationship correctly implies that dwell times approach infinity as $T_{ii}$ approaches 1. 
%  There we found the most success with agglomerative clustering by linearizing this 
%  relationship and clustering on $-log(1 - T_{ii})$.

  \subsection{Choosing Linkage Criteria and the Number of Clusters}\label{section:nclusters}

  In this section, we determine the optimal linkage criteria and number of clusters
  for studying methanol's parameters. We applied the same workflow to make analogous
  decisions for ethylene glycol, urea and acetic acid. In all cases, we found that
  the `ward' linkage criteria resulted in the most useful clustering. We chose to 
  use x number of clusters for each solute, the minimum number of clusters, by our
  analysis, needed to adequately distinguish state dynamics.

  In order to cluster the parameter sets, we need to determine the linkage criteria 
  that will be used to measure the distance between clusters as well as an appropriate
  total number of clusters. In Figure~\ref{fig:linkages}, we show the result of clustering
  with each type of linkage criteria available with the 
  \texttt{sklearn.cluster.AgglomerativeClustering} class of the \texttt{scikit-learn} 
  python package. It is clear that both the `average' and `single' linkage criteria 
  do not sufficiently cluster the data. The `ward' criteria appears to result in the 
  highest number of large population clusters.

  \begin{figure}
  \centering
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{nclusters_ward.pdf}
  \caption{`ward'}\label{fig:nclusters_ward}
  \end{subfigure}
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{nclusters_complete.pdf}
  \caption{`complete'}\label{fig:silhouette_URE}
  \end{subfigure}
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{nclusters_average.pdf}
  \caption{`average'}\label{fig:nclusters_average}
  \end{subfigure}
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{nclusters_single.pdf}
  \caption{`single'}\label{fig:nclusters_single}
  \end{subfigure}
  \caption{In each column we tested a different linkage criteria for clustering
  of the methanol parameters. In each row, we varied the number of total clusters. 
  Both `ward' and `complete' linkage appear to induce clustering
  while the `average' and `single' criteria tend to concentrate parameters in
  only a fraction of the available clusters.}\label{fig:linkages}
  \end{figure}

  We can help to further narrow our decision of linkage criteria by analyzing 
  the parameters of the clusters which it produces. In particular, we should 
  look for well-distinguished clusters on the radial means so that it will be
  easier to connect solute behavior to the membrane pore structure. One way to
  concisely visualize the radial clusters is by their spread. An optimal number
  of clusters should have the same spread in radial means as the unclustered
  data. In Figure~\ref{fig:rspread}, we show that `ward' linkage results in 
  the highest spread in the radials means of the clusters across a range of
  4--30 total clusters. For a low to intermediate number of clusters, the 
  spread of `ward' clusters is significantly larger than of the other linkage
  methods. Therefore, we have chosen to use `ward' clustering.

  We use the Silhouette score in order to aid us in choosing the total number of
  clusters. In Figure~\ref{fig:silhouette_MET}, we plot the Silhouette score as
  a function of the total number of clusters using the `ward' linkage criteria. 
  A Silhouette score of 1 indicates the best clustering possible, values near
  0 indicate overlapping clusters, and negative values indicate samples belong to
  the wrong clusters. For the methanol parameters, the Silhouette score drops off
  quickly with number of clusters before plateauing around 7 total clusters. 
  However, given the low spread in $\mu_r$, it is not clear that using 7 or less 
  clusters will give useful clustering results.

  \begin{figure}
  \centering
  \begin{subfigure}{0.53\textwidth}
  \includegraphics[width=\textwidth]{rspread_nclusters.pdf}
  \caption{}\label{fig:rspread}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{silhouette_MET.pdf}
  \caption{}\label{fig:silhouette_MET}
  \end{subfigure}
  \caption{(a) The `ward' linkage criteria maintains the highest standard 
  deviation in radial means, $\mu_r$, of the clusters for a given total 
  number of clusters as compared to other linkage criteria. (b) The 
  Silhouette score, used to evaluate the quality of clustering with the 
  `ward' distance criteria drops precipitously as a function of the chosen
  total number of clusters before plateauing.}\label{fig:idk}
  \end{figure}

  We incorporate qualitative feedback into the decision on the number of 
  clusters. Despite its high Silhouette score, Figure~\ref{fig:4cluster_state_sequence}
  illustrates that using 4 clusters clearly does not distinguish dynamical 
  modes. We do not see visually acceptable clustering on the example trajectory
  until we use at least twenty total clusters (Figure~\ref{fig:20cluster_state_sequence}).

  \begin{figure}
  \begin{subfigure}{0.9\textwidth}
  \includegraphics[width=1\textwidth]{clustered_traj_MET_ward_4.pdf}
  \caption{4 total clusters}\label{fig:4cluster_state_sequence}
  \end{subfigure}
  \begin{subfigure}{0.9\textwidth}
  \includegraphics[width=1\textwidth]{clustered_traj_MET_ward_12.pdf}
  \caption{12 total clusters}\label{fig:12cluster_state_sequence}
  \end{subfigure}
  \begin{subfigure}{0.9\textwidth}
  \includegraphics[width=1\textwidth]{clustered_traj_MET_ward_20.pdf}
  \caption{20 total clusters}\label{fig:20cluster_state_sequence}
  \end{subfigure}
  \caption{Although the Silhouette score for four total clusters is high, we do
  not see adequate distinction between clusters until we group the parameters
  from the 24 trajectories into 20 total clusters. When we use 12 total clusters,
  the distinction appears acceptable except that the segment from ~50--550 ns
  and from 1200--1500 ns (both colored orange) are clustered together. The radial
  means of these two segments correspond to two very different morphological
  regions of the membrane.}\label{fig:clustered_state_sequences}
  \end{figure}

  \begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{nclusters_msd_MET_ward.pdf}
  \caption{The total MSD predicted by stochastic realizations of the clustered
  model is relatively insensitive to the total number of clusters. All
  predictions are less than the predictions of MD and the unclustered trajectories.
  This behavior is expected due to underestimation of the short dwell time 
  densities as described in Section~\ref{M-section:unclustered_MSD_prediction}
  of the main text.}\label{fig:MET_nclusters} 
  \end{figure}

  This clustering analysis will be finished after some mechanistic studies. It is
  clear that we will need some understanding of mechanisms in order to 
  arrive at a cluster recommendation.

%  \begin{figure}[h]
%  \centering
%  \begin{subfigure}{0.475\textwidth}
%  \includegraphics[width=\textwidth]{nclusters_msd_MET_ward.pdf}
%  \caption{ward}\label{fig:nclusters_msd_MET_ward}
%  \end{subfigure}
%  \begin{subfigure}{0.475\textwidth}
%  \includegraphics[width=\textwidth]{nclusters_msd_MET_complete.pdf}
%  \caption{complete}\label{fig:nclusters_msd_MET_complete}
%  \end{subfigure}
%  \begin{subfigure}{0.475\textwidth}
%  \includegraphics[width=\textwidth]{nclusters_msd_MET_average.pdf}
%  \caption{average}\label{fig:nclusters_msd_MET_average}
%  \end{subfigure}
%  \begin{subfigure}{0.475\textwidth}
%  \includegraphics[width=\textwidth]{nclusters_msd_MET_single.pdf}
%  \caption{single}\label{fig:nclusters_msd_MET_single}
%  \end{subfigure} 
%  \caption{}\label{fig:MET_nclusters} 
%  \end{figure}

  \subsection{Tips for Reliable State Sequence Identification}\label{section:fitting_tips}

  The IHMM is quite powerful on its own, but requires some critical thinking by
  the user in order to ensure maximum effectiveness. One can obtain 
  the best results with some simple checks and, where necessary, improved 
  state sequence initialization.

  The best way to ensure reliable MSD predictions is to ensure proper reproduction
  of the individual MD MSDs by realizations of the IHMM. Over-predictions and 
  under-predictions are usually the result of a poorly identified state sequence. (see
  Figure TBD for example)

  \begin{figure}
  \centering
  \begin{subfigure}{0.35\textwidth}
  \includegraphics[width=\textwidth]{overestimate_ACH_21.pdf}
  \caption{}\label{fig:msd_overestimate}
  \end{subfigure}
  \begin{subfigure}{0.63\textwidth}
  \includegraphics[width=\textwidth]{state_sequence_before_ACH_21.pdf}
  \caption{}\label{fig:state_sequence_before}
  \end{subfigure}
  \begin{subfigure}{0.35\textwidth}
  \includegraphics[width=\textwidth]{msd_improvement_ACH_21.pdf}
  \caption{}\label{fig:msd_improvement}
  \end{subfigure}
  \begin{subfigure}{0.63\textwidth}
  \includegraphics[width=\textwidth]{state_sequence_after_ACH_21.pdf}
  \caption{}\label{fig:state_sequence_after}
  \end{subfigure}
  \caption{(a, b) In some cases, the state segmentation leads to over- or 
  under-prediction of the MSD, usually due to poor initialization of the state
  sequence. In (b), state transitions are too frequent and lead to an
  overprediction of the MD MSD. (c) We can greatly improve MSD predictions by
  giving the IHMM a good guess at the initial state sequence. (d) A better
  initial state segmentation via seeding, gives more reasonable estimates
  of the final state sequence.}\label{fig:improvement_state_sequence}
  \end{figure}

  There are two primary approaches we have implemented in order to improve the final 
  state sequence predictions. Both attempt to aid the initial state assignments 
  algorithmically. First, we always initialize the prior on $\mathbf{c}$ in 
  Equation~\ref{M-eqn:var} of the main text as described in 
  Section~\ref{section:prior_guesses}. In particularly tricky cases, the initial
  state sequence estimate is so poor that qualitative trajectory realizations
  and MSD predictions are irreconcilable with MD. In our second approach, we can
  seed the initial state sequence with a guess made by applying the IHMM to 
  smaller segments of the trajectory.

  \begin{figure}
  \centering
  \begin{subfigure}{0.48\textwidth}
  \includegraphics[width=\textwidth]{seed_ACH21_segment0.pdf}
  \caption{Found 4 Unique States}\label{fig:segment1}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
  \includegraphics[width=\textwidth]{seed_ACH21_segment1.pdf}
  \caption{Found 14 Unique States}\label{fig:segment2}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
  \includegraphics[width=\textwidth]{seed_ACH21_segment2.pdf}
  \caption{Found 13 Unique States}\label{fig:segment3}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
  \includegraphics[width=\textwidth]{seed_ACH21_segment3.pdf}
  \caption{Found 12 Unique States}\label{fig:segment4}
  \end{subfigure}
  \begin{subfigure}{1\textwidth}
  \includegraphics[width=\textwidth]{seed_ACH21_full.pdf}
  \caption{Initial state sequence seeded with states found in (a)--(d)}\label{fig:full}
  \end{subfigure}
  \caption{In (a--d), we show the state sequence resulting from 5 iterations
  of the IHMM on each of the four quarters of the center-of-mass trajectory in
  (e) after splitting it into 4 separate segments. We use the concatenation 
  of the four state sequences as the initial sequence for the IHMM run on the 
  fully trajectory in (e).
  }\label{fig:seed_sequence}
  \end{figure}
